Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-24 analysis/paper/paper.Rmd
updated Study Site

Habitats containing slower water velocities and other attributes that support piscivorous predators are not limited to dams and reservoirs. One such area is the Deadwater Slough in the Salmon River, Idaho which is an unnaturally slow and deep reach of the mainstem river approximately 1.5 kilometers in length and occupying 25.3 acres (Figure \@ref(fig:map)). The slough contains "reservoir-like" conditions which typically favor piscivorous fish predators like Northern Pikeminnow or Smallmouth Bass [@Watkins2015], and further, lacks hydrological and structural features and fish cover (e.g., large wood or substrate, woody debris, bank undercuts) that can provide essential refuge from predation for juvenile fishes. As a result, predation of juvenile Chinook Salmon and other local ESA-listed salmonids (Sockeye Salmon *O. nerka*, redband steelhead *O. mykiss gairdneri*) has been cited as a concern for populations in the Upper Salmon River

@@

The Deadwater Slough is an approximately 1.5 kilometer section of the mainstem Salmon River located roughly 5.8 river kilometers downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River, thereby creating an unnaturally slow and deep section in the river, spanning approximately 30 acres and averaging 68 m width. Both northern pikeminnow *Ptychocheilus oregonensis* and smallmouth bass *Micropterus dolomieu* inhabit the slough which likely provides favorable conditions for their feeding and growth (e.g., reduced water velocity, deep channel, warmer water temperature).

>>>>

Habitats containing slower water velocities and other attributes that support piscivorous predators are not limited to dams and reservoirs. One such area is the Deadwater Slough in the Salmon River, Idaho which is an unnaturally slow and deep reach of the mainstem river approximately 1.5 kilometers in length and occupying 29.5 acres (Figure \@ref(fig:map)). The slough contains "reservoir-like" conditions which typically favor piscivorous fish predators like Northern Pikeminnow or Smallmouth Bass [@Watkins2015], and further, lacks hydrological and structural features and fish cover (e.g., large wood or substrate, woody debris, bank undercuts) that can provide essential refuge from predation for juvenile fishes. As a result, predation of juvenile Chinook Salmon and other local ESA-listed salmonids (Sockeye Salmon *O. nerka*, redband steelhead *O. mykiss gairdneri*) has been cited as a concern for populations in the Upper Salmon River

@@

The Deadwater Slough is located approximately 6 kms downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event that deposited substantial amounts of sediment at the confluence of the Salmon River, forming an alluvial fan and creating an unnaturally slow and deep section of the river resembling a small reservoir.
